### 2024-04-30 Chapters/Chapter1.tex
.

%----------------------------------------------------------------------------------------
Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed. The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period. The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute interval. This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}. As the Navier Stokes Equation (\ref{eq:navierstokes}) shows the variability of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth, and resistance. It states that the change in wind is described by the pressure gradient, the oscillation, the acceleration due to gravity and the resistance from the landscape.
\begin{equation}
    \label{eq:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
\end{equation}
Equation (\ref{eq:navierstokes}) relates the change in wind w.r.t. time ($\frac{\delta \boldsymbol{V}}{\delta t}$) and change in wind w.r.t. position ($\nabla\boldsymbol{V}$) to pressure change w.r.t. position ($\nabla\boldsymbol{P}$). The wind is also influenced by the oscillation, gravity and resistance but the main drive of wind is pressure gradient\cite{uncertainties_in_numerical_weather_predictions}.

Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns\cite{medium_range_3d_weather_forecasting_NN}. These models describe the transition between discretized packages of atmospheric states using partial differential equations based on physical reality. These results are usually published for 3-6 hour intervals.They describe the state over the period and so do not necessarily grasp fluctuations well. These fluctuations would include fluctuations in the wind speed, wind gusts\cite{canNNBeatNWP}.

This thesis looks at how best to predict gust factor based on various factors, using several different data sources, including NWP and observations. NWP forecasts or outputs analysis, traditionally, at intervals of around 3-6 hours and thus, alone, do not capture the variability that is the essence of wind gusts. This is why another model that uses the information produced by the NWP along with observations and landscape information, is needed to try to accurately predict wind gusts. Being able to accurately predict the wind gust is important as it is often the peak wind gusts that will cause failures in structures. A problem that will become increasingly prevalent in the near future\cite{nasa_extreme_weather}.

\section{Background}
The history of numerical weather predictions goes all the way back to the 1920's when Lewis Fry Richardson pioneered the field and tried to produce forecasts. The results were flawed and impractical because of the time it took to calculate was much longer than the forecast period. In the 1950's, with the advent of computers the first operational forecasts emerged. In September of 1954, Rossby and his Stockholm based team produced the first real-time barotropic forecasts. The next year the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton New Jersey, released their first forecasts. These forecasts were for 36 hours at 400, 700 and 900 mb. The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}. The field underwent significant transformation over the next 50 years, and predictions got significantly better.

Currently there has been another transformation in the field of weather prediction driven by artificial intelligence. Interest in AI has come in waves. Some progress is made, then interest dwindles. The interest has been increasing steadily since 2010. Notable work that has driven this wave of interest include increase in computational abilities due to parallel processing in graphical processing units (GPU), convolutional neural networks (CNN), which allowed much faster processing of massive (image) datasets and the availability of large datasets online. It is to be noted that images are grid data with some number of channels. Using CNNs could work on any gridded data where there is some spatial features\cite{canNNBeatNWP}. Since 2018, there has been significant work done in the weather prediction field using AI. In 2018, Dueben and Bauer showed that you can build a NN that can outperform a simple persistence forecast and is competitive with very coarse-resolution atmosphere models of similar complexity for short lead times\cite{dueben2018}. Also in 2018, Scher created a deep convolutional neural network (CNN) to emulate a general circulation model (GCM, a numerical model representing the physical processes), training on the GCM which allows it to emulate the dynamics of the model and maintain stability for much longer than Dueben\cite{scher2018}. These two papers were more proof of concept rather than production ready models to replace NWP. They showed that models based on deep learning might, with further development, compete with standard models in the field.

@@

There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022)\cite{SecondRevolution}. The progression from machine learning methods being an interesting idea in the field of numerical weather predictions, to outperforming the standard NWP has been remarkably quick. Two years ago, machine learning methods were able to predict quickly and in some niche cases outperform traditional models. They were not generally competitive with standard weather models. Now they are competitive. It will be very interesting to watch what the next few years will have in store for the development of machine learning in weather predictions.

@@

In 2004, H. Ágústsson and H. Ólafsson\cite{mean_gust_HA_HO} looked at the variability of gust factor in complex landscapes. They looked at data from automatic weather stations that measure wind at 10 meters above ground. A common definition of wind gust is the maximum average 1 second measured wind speed over a rolling 3 second interval using all measurements over a ten minute span. A common definition of average wind is the average of the measured wind speed over 10 minutes. These are the definitions used both by H. Ágústsson and H. Ólafsson in 2004 as well as this thesis. The data that was studied in 2004 comes from the same source as used in this thesis, but limits itself to a smaller section. They only looked at the years 1999-2001. They looked at three factors and how these three parameters effected the gust factor. These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station. Their main results were these, gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain. The study in 2004 looked at the effect of a dominant point upwind. It did not look at the effects of the landscape more broadly. In this study, landscape upwind is looked at along with landscape downwind. Both the upwind and downwind landscape can influence the gustiness at a certain point \cite{GNP_vidtal}.

>>>>

%----------------------------------------------------------------------------------------
Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed. The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period. The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute intervalf. This thesis uses this definition. This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}. The Navier Stokes Equation (\ref{eq:navierstokes}) shows that the change of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth (the Coriolis force), and frictional force.\cite{uncertainties_in_numerical_weather_predictions}
\begin{equation}
    \label{eq:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
\end{equation}

Traditionally, numerical weather prediction (NWP) systems are used to forecast and analyze weather patterns\cite{medium_range_3d_weather_forecasting_NN}. These models describe the transition between discretized packages of atmospheric states using partial differential equations based on physical reality. These results are usually published every hour, or at courser time intervals for climate simulations. With increasing computer power and efficiency the trend is to output data more often\cite{GNP_vidtal}.They describe the state over the period and so do not necessarily grasp fluctuations well. These fluctuations would include fluctuations in the wind speed, wind gusts\cite{canNNBeatNWP}.

This thesis looks at how best to predict gust factor based on various factors, using several different data sources, including NWP and observations. Being able to accurately predict the wind gust is important as it is often the peak wind gusts that will cause failures in structures. A problem that will become increasingly prevalent in the near future\cite{nasa_extreme_weather}.

\section{Background}
The history of numerical weather predictions goes all the way back to the 1920's when Lewis Fry Richardson pioneered the field and tried to produce forecasts. The results were flawed due to noise in the calculations. ENIAC was built in 1945, it was a general purpose computer that was used, among other things, to make predictions. These predictions took 24 hours to make and were predicting 24 hours into the future. It was a proof of concept but not usable\cite{TheENIACForecastsARecreation}. In the 1950's, with the advent of computers the first operational forecasts emerged. In September of 1954, Rossby and his Stockholm based team produced the first real-time barotropic forecasts. The next year the Joint Numerical Weather Prediction Unit (JNWPU), based in Princeton New Jersey, released their first forecasts. These forecasts were for 36 hours at 400, 700 and 900 mb. The results were inferior to subjective human-based forecasts but showed that such forecasts were feasible and promoted further development in the area \cite{historyNWP}. The field of NWP has taken great strides since then following the development of computer power an efficiency.

Currently there has been another transformation in the field of weather prediction driven by artificial intelligence. Interest in AI has come in waves. Some progress is made, then interest dwindles. The interest has been increasing steadily since 2010. Notable work that has driven this wave of interest include increase in computational abilities due to parallel processing in graphical processing units (GPU), convolutional neural networks (CNN), which allowed much faster processing of massive (image) datasets and the availability of large datasets online. It is to be noted that images are grid data with some number of channels. Using CNNs could work on any gridded data where there are some spatial features\cite{canNNBeatNWP}. Since 2018, there has been significant work done in the weather prediction field using AI. In 2018, Dueben and Bauer showed that you can build a NN that can outperform a simple persistence forecast and is competitive with very coarse-resolution atmosphere models of similar complexity for short lead times\cite{dueben2018}. Also in 2018, Scher created a deep convolutional neural network (CNN) to emulate a general circulation model (GCM, a numerical model representing the physical processes), training on the GCM which allows it to emulate the dynamics of the model and maintain stability for much longer than Dueben\cite{scher2018}. These two papers were more proof of concept rather than production ready models to replace NWP. They showed that models based on deep learning might, with further development, compete with standard models in the field.

@@

There has been a lot of progress made over the last 6 years (since 2018) and especially in the last 2 years (since 2022)\cite{SecondRevolution}. The progression from machine learning methods being an interesting idea in the field of numerical weather predictions, to outperforming the standard NWP has been remarkably quick. Two years ago, machine learning methods were able to predict quickly and in some niche cases outperform traditional models. They were not generally competitive with standard weather models. Now they are competitive. It is worth noting that the training of these large models is based on data from traditional large weather models. It will be very interesting to watch what the next few years will have in store for the development of machine learning in weather predictions.

@@

In 2004, H. Ágústsson and H. Ólafsson\cite{mean_gust_HA_HO} looked at the variability of gust factor in complex landscapes. They looked at data from automatic weather stations that measure wind at 10 meters above ground. The data that was studied in 2004 comes from the same source as used in this thesis, but limits itself to a smaller section. They only looked at the years 1999-2001. They looked at three factors and how these three parameters effected the gust factor. These were $d_m, D, H$, that is direction of wind blowing off a mountain, distance to the mountain and the height of the mountain above the weather station. Their main results were that the gust factor is inversely correlated to the distance from a mountain and correlated to the height of the mountain. The study in 2004 looked at the effect of a dominant point upwind. It did not look at the effects of the landscape more broadly. In this study, landscape upwind is looked at along with landscape downwind. Both the upwind and downwind landscape can influence the gustiness at a certain point \cite{GNP_vidtal}.

### 2024-04-30 Chapters/Chapter2.tex
.

IMO provided 10 minute measurements from \nStationsMin weather stations all around Iceland. The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023 (the year measurements were provided). Of these \nStationsMin stations, \nVedurMin were from IMO Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 7.5 meters above ground. The location of these weather stations can be seen in Figure \ref{fig:aws_map}. The information that is provided by these Automatic Weather Stations (AWS) is presented in two different type of data files, hourly and 10 minute files. The hourly files are summations of the 10 minute files, with the exception that errors, such as nails, still in the 10 minute files should have been removed from the hourly documents. Nails, are sharp increases from the rest of the data and are unrealistic outliers that are considered measurement errors and are discarded. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These measurements were started in the end of the 20th century, when the first AWS stations were installed. More have been added in the over two decades since then. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\begin{figure}
    \includegraphics[scale = 0.75]{Figures/weather_stations_2024-04-09.png}
    \caption{Locations of automatic weather stations in Iceland}
    \label{fig:aws_map}
\end{figure}

@@

\begin{lstlisting}[style = Python, caption = {Filter points over 48 hour interval}, label = code:filter48hr]
    for station in tqdm(stations, total = len(stations)):
        subset_df = vedur_df[station == vedur_df.stod]
        subset_df = subset_df.reset_index(drop = True)

@@

\begin{figure}
    \begin{center}
        \includegraphics[scale = 0.75]{Figures/occurrences_by_year.png}
    \end{center}
    \caption{Measurements that fit filtering by year (from \startDateVedur to 2023).}
    \label{fig:occurrences_by_year}
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics[scale = 0.75]{Figures/station_occurrences_20.png}        
    \end{center}
    \caption{Number of stations that have measurements in a range.}
    \label{fig:occurrences_by_station_20}
\end{figure}

@@

Each request is made by going through all instances of the IMO data after filtering (by mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval). For each such observation in the IMO data, two API calls excluding redundancy are needed. One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times. That is if the observation was at 13:00, CARRA data for noon and 15:00 would have to be queried. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, interpolation was used to get an estimation for the point of the given weather station. The CARRA data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be interpolated to create data for model to be trained on.

@@

\section{Elevation data}
IMO provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid. This file encompasses Iceland and is around 685 MB. A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area. The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes. The Python package Rasterio allowed for quick lookup with it's index and the affine transform. Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.

>>>>

IMO provided 10 minute measurements from \nStationsMin weather stations all around Iceland. The measurements that met the filtering criteria, started in \startDateVedur and ended in 2023 (the year measurements were provided). Of these \nStationsMin stations, \nVedurMin were from IMO Íslands and placed at 10 meters above ground, while the rest (\nVGMin) were from \href{https://www.vegagerdin.is/}{the Icelandic Road and Coastal Administration (IRCA)} and placed at 6-7 meters above ground\cite{vegagerdin_postur}. The location of these weather stations can be seen in Figure \ref{fig:aws_map}. The information that is provided by these Automatic Weather Stations (AWS) is presented in two different type of data files, hourly and 10 minute files. The hourly files are summations of the 10 minute files, with the exception that errors, such as nails, still in the 10 minute files should have been removed from the hourly documents. Nails, are sharp increases from the rest of the data and are unrealistic outliers that are considered measurement errors and are discarded. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another data set), the average wind speed ($f$), the wind gust ($f_g$), the standard deviation of the wind gust, the direction of the wind ($d$) and the standard deviation for the wind direction. These measurement started at the end of the 20th century, when the first AWS stations was installed. More have been added in the following decades. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\begin{figure}
    \centering
    \includegraphics[scale = 0.75]{Figures/weather_stations_2024-04-09.png}
    \caption[Locations of automatic weather stations in Iceland]{Locations of all 412 stations that were looked at in this study. Most of these were from IMO but over a hundred were from IRCA. IMO AWS are placed at 10 meters above ground, while IRCA stations are placed at around 6-7 meters above ground.}
    \label{fig:aws_map}
\end{figure}

@@

\begin{lstlisting}[style = Python, caption = {[Filter points over 48 hour interval] Filter all measured points so to only keep the highest value in any 48 hour interval for a given station. For a given station, the highest wind speed recorded is selected. Then any measurement in a 48 hour interval, centered on the highest wind speed, for that given station, is removed. This is then done iteratively until we are left with an empty dataframe.}, label = code:filter48hr]
    for station in tqdm(stations, total = len(stations)):
        subset_df = vedur_df[station == vedur_df.stod]
        subset_df = subset_df.reset_index(drop = True)

@@

\begin{figure}
    \begin{center}
        \includegraphics[scale = 0.75]{Figures/occurrences_by_year.png}
    \end{center}
    \caption[Measurements that fit filtering by year (from \startDateVedur to 2023).]{The total number of measurements by year after filtering by average wind speed limit of \averageWindSpeedLimit m/s and clearing the neighborhood using filtering method described in code listing. The measurements provided were sparse for years before 2005, as can be seen for 2004. Years before 2005 will thus be ignored.}
    \label{fig:occurrences_by_year}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale = 0.75]{Figures/station_occurrences_20.png}
    \caption[Number of stations that have measurements in a range.]{The number of stations that have a-b measurements represented in filtered data. That is station x may have c measurements represented and $\protect a <= c < b$. Station x will then add to the count of column labeled $\protect [a, b[$.}
    \label{fig:occurrences_by_station_20}
\end{figure}

@@

Each request is made by going through all instances of the IMO data after filtering (by mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval). For each such observation in the IMO data, two API calls excluding redundancy are needed. One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly on one of the three hour interval times. That is if the observation was at 13:00, CARRA data for noon and 15:00 would have to be queried. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, interpolation was used to get an estimation for the point of the given weather station. The CARRA data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be interpolated to create data for model to be trained on.

@@

\subsection{Statistical Analysis of Measurements}
The data gathered from AWS showed inconsistencies due to significant gaps in data coverage in the early 2000s and 90s. This is shown in Figure \ref{fig:measurements_by_year}. It is important to note that Figure \ref{fig:measurements_by_year} and subsequent figures show number of measurements in logarithmic scale. Late in the process of making this thesis, it was discovered that data was missing. Most data points before 2005 had not been provided. Eventually this missing data was provided but the decision was made to exclude that data due to time constraints. Using only data from 2005 to 2023, means that all years cover the entire year. Thus the data was filtered to only include these years. For some reason there were a hundred or so measurements from 1990.

\begin{figure}
    \centering
    \includegraphics[scale = 0.75]{Figures/measurements_by_year.png}
    \caption[Number of measurement data points by year.] {Number of measurement data points by year. The effect of filtering by average windspeed limit shown in the orange overlay. It is important to note that the y-axis scale is logarithmic. If this was not the case, then the filtered values would not be clearly displayed.}
    \label{fig:measurements_by_year}
\end{figure}

It is a good sanity check to look at figures like Figure \ref{fig:measurements_by_year}. If there were massive discrepancies by year, then some data is missing. Another way to look at the distribution of measurement over time is to look at it by month, as in Figure \ref{fig:measurements_by_month}. The total measurements is evenly distributed over the months. As expected higher winds occur during the winter months, with only a few hundred wind measurements reaching \averageWindSpeedLimit.

\begin{figure}
    \centering
    \includegraphics[scale = 0.75]{Figures/measurements_by_month.png}
    \caption[Number of measurement data points by month.]{Number of measurement data points by month. The effect of filtering by average windspeed limit shown in the orange overlay. This filtering shows us what we would expect to see. We can see that we have fewer extreme wind speeds during summer as compared to winter.}
    \label{fig:measurements_by_month}
\end{figure}

Another thing to look at is the number of average wind speed measurements by interval wind speed interval. This is shown in Figure \ref{fig:measurements_by_windspeed}. This figure shows the expected, the number of measurements at higher wind speeds decrease. The orange overlay shows the section that is considered in this thesis.

\begin{figure}
    \centering
    \includegraphics[scale = 0.75]{Figures/measurements_by_windspeed.png}
    \caption[Number of measurement data points by wind speed.]{Number of measurement data points by wind speed. The effect of filtering by average windspeed limit shown in the orange overlay. The filtering obviously just selects the columns to the right and including column $\protect [20, 25[$.}
    \label{fig:measurements_by_windspeed}
\end{figure}

Looking at Figure \ref{fig:measurements_by_windgust} something strange seems to be happening. The orange overlay shows the number of gust values where the wind speed is over \averageWindSpeedLimit m/s and some of the gust values is lower than that. This shows the relevancy of looking at such figures. These data points, where the average wind speed is higher than the gust speed, are filtered out later in the preprocessing pipeline. Note that there are seemingly no missing values for wind speed, wind gust and wind direction, but there is missing standard deviations for these parameters.

\begin{figure}
    \centering
    \includegraphics[scale = 0.75]{Figures/measurements_by_windgust.png}
    \caption[Number of measurement data points by wind gust.]{Number of measurement data points by wind gust. The effect of filtering by average wind speed limit shown in the orange overlay. Here we might expect to see some discrepancies in the columns. That is, the orange overlays should not completely cover all columns. This is because we select by windspeed and not gust. There should be no coverage below the average wind speed limit of \averageWindSpeedLimit m/s.}
    \label{fig:measurements_by_windgust}
\end{figure}

\section{Elevation data}
IMO provided a TIFF file containing the elevation of Iceland on a 20 meter by 20 meter grid. This file encompasses Iceland and is around 685 MB. A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area. The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes. The Python package Rasterio allowed for quick lookup with it's index and the affine transform. Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.

### 2024-04-30 Chapters/Chapter3.tex
.

\begin{figure}[h]
    \caption{A flow chart showing how data sources were combined}
    \label{fig:data_preprocessing_flow_chart}
    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart-2024-04-08.png}
\end{figure}

@@

Figure \ref{fig:error_count} shows that most stations have nails in less that 10\% measurements. The stations that have higher thatn 10\% error, have only few observations that meet the criteria (< 10) and are ignored.

The elevation data comes in a GeoTif file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved interpolation needs to be done. This is done in a similar manner to the interpolation of the CARRA data. The four points bounding the point of interest were used to linearly interpolate the value of the point of interest. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

@@

\begin{table}[h]
    \caption{An example of data structure used with model}
    \label{table:trainDataExample}
    \resizebox{\textwidth}{!}{
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
         Ri\_01 & Ri\_12 & N\_01 & N\_12 & station\_elevation & relative\_corner & PC1 & ...\\\hline
         -1.18e+00 &  2.67e+04 & -8.57e-06 & 6.78e-05 & 3.34e+01 & 2.73e+00 & 1.36e+01 & ... 
    \end{tabular}
    }
\end{table}

Looking at Table \ref{table:trainDataExample} note that the last column shown before ellipsis contains the first principle component in principles component analysis. In total there are \nPCA columns representing the principle component analysis of the elevation data. Also note that the first four columns represent two variables that describe the stability of the air. These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$). They are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}. They are calculated using reanalysis data at two different height levels. Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1. Exactly the same notation is used with the Brunt–Väisälä frequency. The height levels used were 15, 250 and 500 meters above the ground.

@@

Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, $dz$ is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels. Both of these numbers provide some insight about the stability of the air. A lower value for the Richardson number indicates a higher turbulence. A typical range of values could be between 10 and 0.1, with values below 1 indicating significant turbulence\cite{richardson_number_skybrary}. When the square of the Brunt-Väisälä frequency is negative, then the air is unstable (an air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}. These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data. However, including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability. Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded. This the needs to be done for all parameters. The time complexity of this is very high ($2^n$ coalitions)\cite{shapley_information}. Most implementations use some approximations, which still can take a considerable amount of time for models with a high parameter count and many examples.

>>>>

\begin{figure}[h]
    \caption{A flow chart showing how data sources were combined}
    \label{fig:data_preprocessing_flow_chart}
    \includegraphics[scale = 0.5]{Figures/data-preprocessing-flow-chart.png}
\end{figure}

@@

Figure \ref{fig:error_count} shows that most stations have nails in less that 10\% measurements. The stations that have higher than 10\% error, have only few observations that meet the criteria (< 10 measurements) and are ignored.

The elevation data comes in a GeoTIFF file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved interpolation needs to be done. This is done in a similar manner to the interpolation of the CARRA data. The four points bounding the point of interest were used to linearly interpolate the value of the point of interest. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

@@

\begin{table}[h]
    \caption[An example of data structure used with model]{An example of structure of data used to train model. Data points include the derived variables Ri and N, the elevation of the station, direction of wind and relative direction of the wind (that is the direction of the wind relative to center of Iceland), along with some combination of wind speed, pressure and temperature at the different height levels. Finally there are the elevation points around a given station. The example below only shows data derived variables elevation and relative corner.}
    \label{table:trainDataExample}
    \resizebox{\textwidth}{!}{
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
         Ri\_01 & Ri\_12 & N\_01 & N\_12 & station\_elevation & relative\_corner & elevation\_point\_0 & ...\\\hline
         -1.18e+00 &  2.67e+04 & -8.57e-06 & 6.78e-05 & 3.34e+01 & 2.73e+00 & 1.36e+01 & ... 
    \end{tabular}
    }
\end{table}

Looking at Table \ref{table:trainDataExample} note that the last column shown before ellipsis contains the first principle component in principles component analysis. In total there are \nPCA columns representing the principle component analysis of the elevation data. Also note that the first four columns represent two variables that describe the stability of the air. These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$), and are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}. These values are calculated using reanalysis data at three different height levels. Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1. Exactly the same notation is used with the Brunt–Väisälä frequency. The height levels used were 15, 250 and 500 meters above the ground and the three different sets of height levels were (15 m, 250 m), (250 m, 500 m) and (15 m, 500 m).

@@

Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, $dz$ is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels. Both of these numbers provide some insight about the stability of the air. A lower value for the Richardson number indicates a higher turbulence. A typical range of values could be between 0.1 and 10, with values below 1 indicating significant turbulence\cite{richardson_number_skybrary}. When the square of the Brunt-Väisälä frequency is negative, then the air is unstable (an air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}. These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data. However, including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability. Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded. This   needs to be done for all parameters. The time complexity of this is very high ($2^n$ coalitions)\cite{shapley_information}. Most implementations use some approximations, which still can take a considerable amount of time for models with a high parameter count and many examples.
